---
title: "Parent–child interactions as mediators of adolescent outcomes"
author: Lily X. Gates (Van Hout)
format: html
editor: visual
---

# Load required packages

```{r}
# Load Packages
library(ggplot2) # For graphing
library(jtools) # For APA format
library(tidyverse) # For data management, includes dplyr

# Load PROCESS

# Via direct path
#source("~/UMD 2024-2025/Summer 2025/PSYC 489K/Week 3 - Mediation (Part II)/Lecture Materials/Lecture Code/process.R")

# Assume "process.R" is in cwd
source("./process.R")
```

# Data and data prepping

## Read in and summarize data

```{r}
# Contains outcome variables (internalizing vs. externalizing psycopathology symptoms) and PCI data (mediators)
lab_data <- read.csv("lab_data.csv")
#summary(lab_data)
#head(lab_data)

# Contains negative life experiences (NLE) from surveys
nle_data <- read.csv("NLE_data.csv")
#head(nle_data)

# Contains variable names for...
  # Lab experiments
lab_labels <- read.csv("lab_data_var_labels.csv")
#head(lab_labels)
  # NLE
nle_labels <- read.csv("NLE_template_spreadsheet.csv")
#head(nle_labels)
```

## Data Cleaning

**Merging Data**

```{r}
# MERGING
# lab_data: ID is primary key
# nle_data: famid is primary key
# lab_labels: lab_data_var is primary key
# nle_labels: item_code is primary key

# Create merged dataframe of lab data (incl outcomes) and NLE survey data
lab_nle_data <- lab_data %>%
  inner_join(nle_data,
             by = c("ID" = "famid"))

#head(lab_nle_data)
```

**Renaming Columns**

```{r}
#RENAMING COLUMNS

#colnames(lab_nle_data)

full_df <- lab_nle_data %>%
  rename(
    childs_sex = sex,
    internal_psyc = INTERNAL,
    external_psyc = EXTERNAL,
  )

#colnames(full_df)
```

**Dropping Columns**

```{r}
# The columns to drop
# Columns to drop: index ("X"), name of coder ("o1coder")
drop <- c("X", "o1coder")

# Filtered df when removing "drop" columns
filtered_df <- full_df[,!(names(full_df) %in% drop)]

#colnames(filtered_df)
```

**Convert to Appropriate Columns to Numeric**

```{r}
char_col <- c("ID", "childs_sex", "o1rpinscor", "o1rpinscom", "o1rpothprob")

# Convert all columns to numeric except char_col

filtered_df <- filtered_df %>%
  mutate(across(.cols = -all_of(char_col), .fns = ~as.numeric(.)))
```

```{r}
# Checking data type
#glimpse(filtered_df)
#sapply(filtered_df, class)

# Identify columns that are NOT numeric and NOT in char_col
cols_to_convert <- setdiff(names(filtered_df)[!sapply(filtered_df, is.numeric)], char_col)
print(cols_to_convert)

# Convert only those columns
filtered_df[cols_to_convert] <- lapply(filtered_df[cols_to_convert], as.numeric)

```

# Analyze Participants from Data Set

Children and their mothers residing in the Seattle, Washington area were recruited to participate in a longitudinal study conducted at the University of Washington. The initial phase of data collection began in 2008 under the supervision of Dr. Liliana Lengua, during which parent–child interactions were systematically observed and coded while the children were in early childhood. Subsequently, the study was continued by Dr. Katie McLaughlin, who conducted follow-up interviews with both children and their mothers to assess internalizing and externalizing psychopathology experienced by the children during adolescence.

## Demographics

The sample consisted of 227 children (*N* = 227).

The mean age of the total sample was 3.05 years (*M* = 3.05, *SD* = 0.07, range = 2.95–3.34).

Among them, 114 were male (*n* = 114), with a mean age of 3.05 years (*M* = 3.05, *SD* = 0.07, range = 2.95–3.32), and 113 were female (*n* = 113), with a mean age of 3.05 years (*M* = 3.05, *SD* = 0.07, range = 2.96–3.34).

```{r}
# TOTAL SAMPLE SIZE
N <- sum(!is.na(filtered_df$c1ageyr))
print(N)

# MIN, MAX, AVERAGE OF CHILD AGE
# Minimum
min(filtered_df$c1ageyr, na.rm = TRUE)

# Maximum
max(filtered_df$c1ageyr, na.rm = TRUE)

# Mean (average)
mean(filtered_df$c1ageyr, na.rm = TRUE)

# Standard deviation
sd(filtered_df$c1ageyr, na.rm = TRUE)

# SUBSET BY GENDER OF CHILD
# Subset for males
males <- filtered_df[filtered_df$childs_sex == "Male", ]

# Calculate sample size, mean, sd, and range for c1ageyr in males
n_males <- sum(!is.na(males$c1ageyr))
mean_males <- mean(males$c1ageyr, na.rm = TRUE)
sd_males <- sd(males$c1ageyr, na.rm = TRUE)
range_males <- range(males$c1ageyr, na.rm = TRUE)

cat("Male children (n =", n_males, "): Mean age =", round(mean_males, 2), 
    "years, SD =", round(sd_males, 2), ", Range =", round(range_males[1], 2), 
    "-", round(range_males[2], 2), "years.\n")

# Subset for females
females <- filtered_df[filtered_df$childs_sex == "Female", ]

# Calculate sample size, mean, sd, and range for c1ageyr in females
n_females <- sum(!is.na(females$c1ageyr))
mean_females <- mean(females$c1ageyr, na.rm = TRUE)
sd_females <- sd(females$c1ageyr, na.rm = TRUE)
range_females <- range(females$c1ageyr, na.rm = TRUE)

cat("Female children (n =", n_females, "): Mean age =", round(mean_females, 2), 
    "years, SD =", round(sd_females, 2), ", Range =", round(range_females[1], 2), 
    "-", round(range_females[2], 2), "years.\n")


```

## Socioeconomic Status: income-to-needs (ITN)

```{r}
# 'ITN'

# Mean
mean_value <- mean(filtered_df$ITN, na.rm = TRUE)

# Range (min and max)
range_values <- range(filtered_df$ITN, na.rm = TRUE)

# Standard Deviation
sd_value <- sd(filtered_df$ITN, na.rm = TRUE)

# Print results
cat("Mean:", mean_value, "\n")
cat("Range:", range_values[1], "to", range_values[2], "\n")
cat("Standard Deviation:", sd_value, "\n")
```

# Multiple Mediation

**Predictors:** Negative Life Events (NLE)

**Parallel Mediation Equations for Parent-Child Interactions**

-   **Warmth:** M1 = d_m1 + a_1\*X

-   **Rejection:** M2 = d_m2 + a_2\*X

-   **Structure:** M3 = d_m3 + a_3\*X

-   **Interactiveness:** M4 = d_m4 + a_4\*X

**Outcomes:** Internal Psychopathology and External Psychopathology

Y = d_Y + c'\*X + b_1\*M_1 + b_2\*M_2 + b_3\*M_3 + b_4\*M_4

When controlling for *all other mediators*, the specific indirect effect of *X* on *Y* through...

-   M_1 is *a_1*\**b_1*

-   M_2 is *a_2*\**b_2*

-   M_3 is *a_3*\**b_3*

-   M_4 is *a_4*\**b_4*

The **total indirect effect of *X* on *Y***

*a_1*\**b_1* + *a_2*\**b_2* + *a_3*\**b_3 + a_4*\**b_4*

Adding this to *c'* will gives the total effect, which is *Y* regressed on *X* alone.\
process(data = df, x = "X", m = c("M1","M2","M3","M4"), y = "Y", model = 4).

Note: model = 4 applies to simple mediation models and parallel mediation models.

# **Comparing Indirect Effects in PROCESS**

If the indirect effects are the [same]{.underline} sign, simply, use the [raw]{.underline} difference of the indirect effects (contrast = 1)

If the indirect effects are of [different]{.underline} signs, take the difference of [absolute values]{.underline} or compute the sum of the indirect effects (contrast = 2)

# Organizing Predictors (NLE)

-   "response_0_to_3", "did_not_happen_0", "not_upsetting_1", "a_little_upsetting_2", "very_upsetting_3"

NOTE: TC means "the child"

**Transition**

-   m1le019 – A close friend of TC’s moved away

-   m1le022 – TC moved or there was a change in TC’s living situation

-   m1le023 – You had a new baby

-   m1le024 – TC was separated from his/her family for a long time

-   m1le025 – You divorced or separated in the last year

-   m1le026 – You or TC’s other parent remarried in the last year

-   m1le027 – You had a new \[boyfriend/girlfriend\] move into the family home

-   m1le028 – TC changed schools

**Death**

-   m1le018 – A relative or close family friend died

-   m1le029 – TC had a family pet pass on (die)

**Illness / Injury**

-   m1le002 – TC’s close friends had troubles, an illness or injury

-   m1le003 – TC suffered a serious illness or injury

-   m1le004 – TC’s brother or sister suffered serious illness or injury

-   m1le008 – You or TC’s other parent suffered serious illness or injury

**Finances**

-   m1le009 – You talked about money troubles in TC’s presence

-   m1le017 – You or TC’s other parent lost a job

**Psychological Stress**

-   m1le005 – TC’s brother or sister was very angry or upset

-   m1le010 – Your relatives said bad things about you in TC’s presence

-   m1le011 – You fought or argued with relatives in TC’s presence

-   m1le012 – Neighbors said bad things about you to TC

-   m1le020 – You or TC’s other parent acted worried or sad in TC’s presence

-   m1le021 – TC’s friends teased him/her or were mean to him/her

**Violence / Conflict Exposure**

-   m1le006 – You and your partner hit each other in TC’s presence

-   m1le007 – People in TC’s family hit each other

**Behavioral / Legal Disruption**

-   m1le001 – TC’s brother or sister had troubles with the law

-   m1le015 – You or TC’s other parent forgot to do important things for TC

-   m1le016 – You or TC’s other parent was arrested or sent to jail

**Sorting NLE into Vectors by Category**

Using "Theoretical Grouping", but could potentially use factorial analysis to analyze the latent variables. For example, Exploratory Data Analysis or Confirmatory Factor Analysis (Structural Equation Modeling).

```{r}
# Define all groups in a named list
nle_groups <- list(
  trans = c("m1le019", "m1le022", "m1le023", "m1le024", "m1le025", "m1le026", "m1le027", "m1le028"),
  death = c("m1le018", "m1le029"),
  ill_injury = c("m1le002", "m1le003", "m1le004", "m1le008"),
  finances = c("m1le009", "m1le017"),
  psyc_stress = c("m1le005", "m1le010", "m1le011", "m1le012", "m1le020", "m1le021"),
  violence = c("m1le006", "m1le007"),
  behav_legal = c("m1le001", "m1le015", "m1le016")
)

```

## Predictor of NLE using "Total Severity"

This will use "Total Severity," by using sum of ratings. Another approach could be "Total Number of Events" by summing any instances of non-zeros.

**Creating Columns with NLE Averages by Category**

```{r}
# Loop through the list and compute means
for(group_name in names(nle_groups)) {
  
  # Select the appropriate columns with the same name in the df
  cols <- nle_groups[[group_name]]
  
  # Check columns exist in df to avoid errors
  cols <- cols[cols %in% names(filtered_df)]
  
  # Rename the column in a format of nle_[group_name]_mean
  new_col_name <- paste0("nle_", group_name, "_mean")
  
  # Make the column contain an overall average of NLE
  filtered_df[[new_col_name]] <- rowMeans(filtered_df[cols], na.rm = TRUE)
}

#colnames(filtered_df)
```

**Create a Column with Average of All NLE**

```{r}
# Extract group names
group_names <- names(nle_groups)

# Find columns that match the naming pattern in the df
nle_mean_cols <- paste0("nle_", group_names, "_mean")
nle_mean_cols <- nle_mean_cols[nle_mean_cols %in% names(filtered_df)]

# Add column that contains average of all NLE; ignore NAs
filtered_df$nle_total_mean <- rowMeans(filtered_df[nle_mean_cols], na.rm = TRUE)

#filtered_df$nle_total_mean
```

## Predictor of NLE using "Total Number of Events"

**Calculated by summing any instances of non-zeros for NLE categories.**

```{r}
# Loop through the list and compute total number of events
for(group_name in names(nle_groups)) {
  
  # Select the appropriate columns with the same name in the df
  cols <- nle_groups[[group_name]]
  
  # Check columns exist in df to avoid errors
  cols <- cols[cols %in% names(filtered_df)]
  
  # Rename the column in a format of nle_[group_name]_total_events
  new_col_name <- paste0("nle_", group_name, "_total_events")
  
  # Make the column contain total number of events (count non-zeros); ignore NAs
  filtered_df[[new_col_name]] <- rowSums(filtered_df[cols] != 0, na.rm = TRUE)

}
```

**Add column with total sum of all event occurrences**

```{r}
# Extract group names
group_names <- names(nle_groups)

# Find columns that match the naming pattern
nle_total_event_cols <- paste0("nle_", group_names, "_total_events")
nle_total_event_cols <- nle_total_event_cols[nle_total_event_cols %in% names(filtered_df)]

# Add column that contains total sum of events; ignore NAs
filtered_df$nle_all_total_events <- rowSums(filtered_df[nle_total_event_cols], na.rm = TRUE)
```

## Predictors Renamed

```{r}
# Predictors Renamed
NLE_num_events <- filtered_df$nle_all_total_events
NLE_severity <- filtered_df$nle_total_mean
```

### Analyzing NLE number of events and severity

Participants reported an average of 5.69 (*SD* = 4.30) negative life events (NLEs), with a range from 0 to 27 events, reflecting considerable variability in exposure.

The mean severity rating of these events was 0.34 (*SD* = 0.28), ranging from 0 to 1.94. Severity scores represent the average distress rating per event on a 0–3 scale, where 0 indicates the event did not occur. Since mothers rated 29 possible events, these findings suggest that, on average, participants experienced several distressing events with mild to moderate perceived impact.

```{r}
# For number of NLE events
mean_events <- mean(filtered_df$nle_all_total_events, na.rm = TRUE)
sd_events <- sd(filtered_df$nle_all_total_events, na.rm = TRUE)
range_events <- range(filtered_df$nle_all_total_events, na.rm = TRUE)

# For NLE severity
mean_severity <- mean(filtered_df$nle_total_mean, na.rm = TRUE)
sd_severity <- sd(filtered_df$nle_total_mean, na.rm = TRUE)
range_severity <- range(filtered_df$nle_total_mean, na.rm = TRUE)

# Print results
cat("NLE Number of Events - Mean:", mean_events, "SD:", sd_events, "Range:", range_events[1], "to", range_events[2], "\n")
cat("NLE Severity - Mean:", mean_severity, "SD:", sd_severity, "Range:", range_severity[1], "to", range_severity[2], "\n")

```

# Organizing Mediators

```{r}
# Extract column names for relevant mediators
pci_vars <- c(
  "o1mpaa",    # Mom Positive Affect
  "o1mnaa",    # Mom Negative Affect
  "o1minta",   # Mom Interactiveness
  "o1mrata",   # Respect for Autonomy
  "o1mnca",    # Negative Control
  "o1mgsa",    # Guidance & Structuring
  "o1mlta",    # Limit Setting
  "o1mrespa",  # Responsiveness (corrected)
  "o1ccoa",    # Child Compliance
  "o1cnaa",    # Child Negative Affect
  "o1cpaa",    # Child Positive Affect
  "o1mncar",   # Negative Control Reversed
  "o1mscaf",   # Scaffolding
  "o1mwrm"     # Warmth
)
```

**User-Friendly Description of Mediator Variables**

```{r}
# Define a named vector for clearer output column names, variable name mapping
pci_domains <- c(
  mpaa  = "mom_pos_affect",
  mnaa  = "mom_neg_affect",
  minta = "mom_interactiveness",
  mrata = "mom_respect_autonomy",
  mnca  = "mom_neg_control",
  mgsa  = "mom_guidance_structuring",
  mlta  = "mom_limit_setting",
  mrespa = "mom_responsiveness",  # <-- corrected from mrspa
  ccoa  = "child_compliance",
  cnaa  = "child_neg_affect",
  cpaa  = "child_pos_affect",
  mncar = "mom_neg_control_reversed",
  mscaf = "mom_scaffolding",
  mwrm  = "mom_warmth"
)

# Loop to create new pci_ columns for GO1 only
# Convert to numeric
for (code in names(pci_domains)) {
  var_name <- paste0("o1", code)
  new_col <- paste0("pci_", pci_domains[[code]])
# If there is missing data, will show warning
  if (var_name %in% names(filtered_df)) {
    filtered_df[[new_col]] <- as.numeric(filtered_df[[var_name]])
  } else {
    warning(paste("Missing column:", var_name))
    filtered_df[[new_col]] <- NA
  }
}

#colnames(filtered_df)
#pci_domains
```

**Sort PCI by Category**

```{r}
# Warmth
  # Latent: Positive Affect, Responsiveness
pci_warmth <- c("o1mpaa", "o1mrespa")

# Rejection
  # Latent: Negativity, Negative Control
pci_rejection <- c("o1mnaa", "o1mnca")

# Structure
  # Latent: Guidance, Autonomy, Compliance
pci_structure <- c("o1mgsa", "o1mrata", "o1ccoa")

# Interactiveness
  # Observed (standalone): Interactiveness score
pci_interactiveness <- "o1minta"

```

**Creating Columns for PCIs**

```{r}
# Compute average for warmth, rejection, structure
filtered_df$pci_warmth <- rowMeans(filtered_df[pci_warmth], na.rm = TRUE)
filtered_df$pci_rejection <- rowMeans(filtered_df[pci_rejection], na.rm = TRUE)
filtered_df$pci_structure <- rowMeans(filtered_df[pci_structure], na.rm = TRUE)

# Note: Not computing average because stand alone variable
filtered_df$pci_interactiveness <- as.numeric(filtered_df[[pci_interactiveness]])
```

# **Organizing Covariate(s)**

**Potential Covariates:** Sex of the child (childs_sex), age of child (c1ageyr), mother's years of education (Mom_edu_cont), socioeconomic status/income-to-needs ratio (ITN)

**Coding for Child's Sex**

```{r}
# Find the unique variables in the childs_sex column
#unique(filtered_df$childs_sex)

# Convert childs_sex into factor variable

filtered_df$childs_sex <- factor(filtered_df$childs_sex, levels = c("Male", "Female"))

# Create logical binary coded column for child's sex
  # Female = 1, Male = 0
filtered_df$childs_sex_num <- ifelse(filtered_df$childs_sex == "Female", 1, 0)

# Convert logical to numeric
# PROCESS cannot handle factors or non-numeric (incl logical)
filtered_df$childs_sex_num <- as.numeric(filtered_df$childs_sex_num)
```

#### **Download 'filtered_df' as a .CSV**

```{r}
library(readr) # Needed to download an r object as CSV
write_csv(filtered_df, "complete_data_set.csv")
```

# Parallel Mediation Model

## **Setting Up Parallel Mediation in PROCESS**

Checking data types and any missing values

NOTE: Some cases with missing data were deleted. The number of deleted cases was: 11

```{r}
# model = 4 is for simple or parallel mediation
# contrast = 2 is for difference in absolute values
# total = 1 shows total effect
# progress = 0 turns off bootstrap progress bar
# seed = 42 makes results reproducible
# boot = number of bootstrap samples
```

```{r}
# PREPPING - Checking data types and any missing values

# All varibales being used for model
vars <- c(
  "nle_total_mean",
  "nle_all_total_events",
  "pci_mom_pos_affect",
  "pci_mom_neg_affect",
  "pci_mom_interactiveness",
  "pci_mom_respect_autonomy",
  "pci_mom_neg_control",
  "pci_mom_guidance_structuring",
  "pci_mom_limit_setting",
  "pci_mom_responsiveness",
  "pci_child_compliance",
  "pci_child_neg_affect",
  "pci_child_pos_affect",
  "pci_mom_neg_control_reversed",
  "pci_mom_scaffolding",
  "pci_mom_warmth",
  "pci_warmth",
  "pci_rejection",
  "pci_structure",
  "pci_interactiveness",
  "childs_sex_num"
)


# Check class types
#sapply(filtered_df[, vars], class)

# Check complete columns
complete_cases <- filtered_df[complete.cases(filtered_df[, vars]), ]
nrow(complete_cases)

# Checking missing values
colSums(is.na(filtered_df[, vars]))

# Rows with any NA in the vars columns
missing_rows <- filtered_df[!complete.cases(filtered_df[, vars]), ]

# View missing rows
# Note: potentially because of audio/visual problems unable to code for PCIs
#missing_rows

```

## **Running the Models**

### **Internal Psychopathology**

```{r}
# RUNNING - Internal Psychopathology

# Model 1 - Total Severity: Using nle_total_mean as the predictor for internal psychopathology
process(
  data = filtered_df,
  x = "nle_total_mean",
  y = "internal_psyc",
  m = c("pci_warmth", "pci_rejection", "pci_structure", "pci_interactiveness"),
  cov = c("childs_sex_num", "c1ageyr", "ITN"),
  model = 4,
  contrast = 2,
  seed = 42,
  boot = 5000,
  total = 1,
  progress = 0,
)

# Model 2 - Total Number of Events: Using nle_all_total_events as the predictor for internal psychopathology
process(
  data = filtered_df,
  x = "nle_all_total_events",
  y = "internal_psyc",
  m = c("pci_warmth", "pci_rejection", "pci_structure", "pci_interactiveness"),
  cov = c("childs_sex_num", "c1ageyr", "ITN"),
  model = 4,
  contrast = 2,
  seed = 42,
  boot = 5000,
  total = 1,
  progress = 0
)
```

### **External Psychopathology**

```{r}
# RUNNING: External Psychopathology

# Model 3 - Total Severity & External Psycopathology: Using nle_total_mean as the predictor for external psychopathology
process(
  data = filtered_df,
  x = "nle_total_mean",
  y = "external_psyc",
  m = c("pci_warmth", "pci_rejection", "pci_structure", "pci_interactiveness"),
  cov = c("childs_sex_num", "c1ageyr", "ITN"),
  model = 4,
  contrast = 2,
  seed = 42,
  boot = 5000,
  total = 1,
  progress = 0,
)

# Model 4 - Total Number of Events & External Psycopathology: Using nle_all_total_events as the predictor for external psychopathology
process(
  data = filtered_df,
  x = "nle_all_total_events",
  y = "external_psyc",
  m = c("pci_warmth", "pci_rejection", "pci_structure", "pci_interactiveness"),
  cov = c("childs_sex_num", "c1ageyr", "ITN"),
  model = 4,
  contrast = 2,
  seed = 42,
  boot = 5000,
  total = 1,
  progress = 0
)
```
